Replicating Blair and Sambanis: Forecasting Civil Wars - JCR 2020

### 2020-04-30 issues.Rmd
update issues note

@@

While unorthodox, the approach does work. It produces predictions that are within the 0 to 1 interval. The concern is that the settings work well only for the escalation model, and only for the particular test set at hand. Specifically we wonder if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.

The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model and a modified version which uses more conventional random forest parameter settings (classification, 5000 trees, and default settings otherwise). The first and second columns are average and SD AUC-ROC obtained through repeated cross-validation on the training data set. Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e. splitting the original training data into equally-sized new training and validation sets. This is repeated 21 times for a total of 42 OOS performance samples for each model. The last column shows the OOS performance on the original test set. The first value in this column corresponds to the base escalation model results reported Table 1 in the B&S paper when *not* using smoothed ROC curves.

```{r}
tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo"
), Avg_CV_ROC_AUC = c(0.698772142673124, 0.643982732304452), 
    SD_CV_ROC_AUC = c(0.136834250535024, 0.104102374345325), 
    Test_ROC_AUC = c(0.786872878159185, 0.583391239320482)), row.names = c(NA, 
-2L), class = c("tbl_df", "tbl", "data.frame"))

@@

Two thigs stand out:

- The escalation model has better performance in the test data than in the training CV samples. This is consistent with hyperparameter values that were (over-)fit to the test data. 
- Although the original escalation model outperforms the modified model in the training CV resamples (p < 0.05), the difference in means is much less pronounced than in the test data.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

While unorthodox, the approach does work. It produces predictions that are within the 0 to 1 interval. The concern is that the settings work well only for the escalation model, and only for the particular test set at hand. Specifically, we wonder (1) if the hyperparameter settings only work well for the particular test set chosen, and (2) if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.

## Are the RF hyperparameters overfit to the test set?

Does the base specification's good performance in the test set generalize?

One initial piece of evidence is already available from Table 1. In addition to the test set starting 2008, B&S also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1). For the 1 month models, when we compare the base specification performance of each model to the performance in the alternate test sets, the AUC-ROC increases in 2 (of 15) cases, is the same in 0 cases, and decreases in 13 cases. For the 6 month models, it increases in 1 case, is the same in 3 cases, and decreases in 11 cases. Altering the test set thus generally shows reduced performance. 
The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model, a RF model with default hyperparameter settings, and a tuned RF model with 10,000 trees and otherwise the default settings. The first set of results are out of sampel results from repeated cross-validations performed on the training data. We show the average AUC-ROC, it's lower and upper 95\% CI, obtained via bootstrapping, and the standard deviation of the distribution of resampled AUC-ROC values.

Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e. splitting the original training data into equally-sized new training and validation sets. This is repeated 21 times for a total of 42 OOS performance samples for each model. The last column shows the OOS performance on the original test set. The first value in this column corresponds to the base escalation model results reported Table 1 in the B&S paper when *not* using smoothed ROC curves.

```{r}
tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo", 
"Tuned Escalation, 1mo"), Avg_CV_ROC_AUC = c(0.677707841149067, 
0.619719616721681, 0.671630392456263), ci_lower = c(0.634101138978023, 
0.5835673668958, 0.634713167892701), ci_upper = c(0.722598486545989, 
0.65584632967561, 0.706310478133188), SD_CV_ROC_AUC = c(0.147542574525528, 
0.122337186539073, 0.117973980721395), Test_ROC_AUC = c(0.783352194140576, 
0.586147564308735, 0.582927991423296)), row.names = c(NA, -3L
), class = c("tbl_df", "tbl", "data.frame"))

@@

The results show that:

- Both of the alternative RF models are able to achieve roughly similar OOS performance in the training data split. *T*-tests comparing model 1 to the other models fail to reject the null hypothesis at a 95\% confidence level. (A *t*-test comparing model 1 and 2 average AUC-ROC just barely fails to reject the null hypothesis (*p* slightly above 0.05).)
- Only the base escalation model is able to achive good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.

This suggests that the base RF specifications are (over-)fit to the test data.
